
\chapter{Intuitionistic Logic and The Curry--Howard Isomorphism}

Unlike classical logic, which is often extended through modal operators or second-order quantifiers, intuitionistic logic takes a different path, not because it adds power, but because it restricts classical reasoning. Classical logic freely allows non-constructive arguments, proofs that assert the existence of an object without providing a way to explicitly construct it. Intuitionistic logic, in contrast, is designed to formalize constructive reasoning, where existence is meaningful only if a specific example can be provided.

\section{\centering Intuitionistic Logic}
To motivate the introduction of Intuitionistic Logic, we present a theorem and two different styles of proof \cite{openlogicprojectintuitionisticlogic2025}:
\begin{theorem}
There are irrational numbers $a$ and $b$ such that $a^b$ is rational.
\end{theorem}
\begin{proof}
  Let $ a, b = \sqrt{2} $
  \begin{itemize}
  \item If $\sqrt{2}^{\sqrt{2}}$ is rational, we are done.
  \item Otherwise, $\sqrt{2}^{\sqrt{2}}$ is irrational and we can consider \( a = \sqrt{2}^{\sqrt{2}} \text{ and } b = \sqrt{2} \). In which case:
    \[
      a^b = \left( \sqrt{2}^{\sqrt{2}} \right)^{\sqrt{2}} = \sqrt{2}^{\sqrt{2}\cdot \sqrt{2}}
      = \sqrt{2}^2 = 2,
    \]
    which is rational so $a$ and $b$ must exist.
  \end{itemize}
  \end{proof}

  \medskip
  This is a perfectly valid proof. However, it is somewhat unsatisfying, since it proves existence without exhibiting a specific pair. A more constructive proof is as follows
\begin{proof}
  \medskip
  Take $a = \sqrt{3}$ and $b = \log_3 4$. Then
  \[
    a^b = \left(\sqrt{3}\right)^{\log_3 4}
    = 3^{\tfrac{1}{2}\log_3 4}
    = \left(3^{\log_3 4}\right)^{1/2}
    = 4^{1/2} = 2,
  \]
  which is rational.
\end{proof}

\subsubsection{\centering Syntax}
The syntax for intuitionistic logic is the same as that of propositional logic. Where $\lor, \land, \to$ are used as binary connectives to compose logical statements and $\lnot$ is used to negate such statements.

\begin{definition}
The set of formulas of propositional intuitionistic logic is defined as follows:
\begin{itemize}
    \item $\bot$ is an atomic formula.
    \item Every propositional variable $p_i$ is a formula.
    \item If $\varphi$ and $\psi$ are formulas, then $(\varphi \wedge \psi)$ is a formula.
    \item If $\varphi$ and $\psi$ are formulas, then $(\varphi \vee \psi)$ is a formula.
    \item If $\varphi$ and $\psi$ are formulas, then $(\varphi \to \psi)$ is a formula.
\end{itemize}

\medskip
In addition to the connectives introduced above, the following symbols define the following symbols to simplify notation:
\begin{itemize}
    \item $\lnot \varphi$ abbreviates $(\varphi \to \bot)$.
    \item $\varphi \leftrightarrow \psi$ abbreviates $(\varphi \to \psi) \wedge (\psi \to \varphi)$.
\end{itemize}
\end{definition}

\subsubsection{\centering The Brouwer--Heyting--Kolmogorov Interpretation}

The Brouwer--Heyting--Kolmogorov interpretation provides an informal, constructive understanding of intuitionistic connectives. It is based on the notion of a construction, which can be regarded as a constructive proof. The BHK interpretation assigns meanings to the intuitionistic connectives as
follows:

\begin{itemize}
  \item We assume that we know what constitutes a construction for an atomic proposition.

  \item A construction of
  $\varphi_1 \wedge \varphi_2$ is a pair $\langle M_1, M_2\rangle$, where
  $M_1$ is a construction of $\varphi_1$ and $M_2$ is a construction of
  $\varphi_2$.

  \item A construction of
  $\varphi_1 \vee \varphi_2$ is a pair $\langle s, M\rangle$, where $s = 1$
  and $M$ is a construction of $\varphi_1$, or $s = 2$ and $M$ is a
  construction of $\varphi_2$.

  \item A construction of
  $\varphi \to \psi$ is a function that converts any construction of
  $\varphi$ into a construction of $\psi$.

  \item There is no construction for $\bot$; it
  represents absurdity.

  \item Defined as $\varphi \to \bot$.
  Thus, a construction of $\neg\varphi$ is a function that converts a construction of $\varphi$ into a construction of $\bot$.
\end{itemize}

\begin{remark}
  Generally speaking, $\neg\varphi$
  means any hypothetical construction $\varphi$ is impossible of $\varphi$ yields absurdity $\bot$.
  
\end{remark}

\section{\centering Natural Deduction}

The syntax that to create formulas is not sufficient, we need to pair it with a system that enables the deduction of other formulas from a set of given ones.

\newcommand{\Fm}{\mathcal{F}}
\newcommand{\Rules}{\mathcal{R}}
\newcommand{\Axioms}{\mathcal{A}}
\newcommand{\Cn}{\mathrm{Cn}}

\begin{definition} A deductive system is a pair
\(
\mathcal{D} = (\Axioms, \Rules),
\)
where:
\begin{itemize}
    \item $\Axioms \subseteq \Fm$ is a set of axioms and $\Fm$ is the set of formulas.
    \item $\Rules$ is a set of inference rules used to derive new formulas from the already known ones.
\end{itemize}
A derivation or proof in $\mathcal{D}$ from a set of premises $\Gamma \subseteq \Fm$ is a finite sequence of formulas
\(
\varphi_1, \dots, \varphi_n
\)
such that each $\varphi_i$ is either:
\begin{itemize}
    \item an axiom ($\varphi_i \in \Axioms$).
    \item a member of $\Gamma$, or obtained from previous formulas in the sequence by an inference rule in $\Rules$.
\end{itemize}
We write
\(
\Gamma \vdash \varphi
\)
if there exists a derivation of $\varphi$ from $\Gamma$ in the deductive system $\mathcal{D}$.
\end{definition}
Natural deduction is a deduction system within intuitionistic logic that uses constructive reasoning to derive conclusions. This structured proofs emphasize the existence of explicit constructions rather than abstract truth values.

The rules of natural deduction for intuitionistic propositional logic can be presented as follows:
\subsubsection{\centering Introduction / Elimination Rules}

\medskip

\[
\begin{array}{@{}l@{\qquad}l@{}}
\inferrule*[right=\textsc{IntroConj}]
  { \phi \\ \psi }
  { \phi \land \psi }
&
\inferrule*[right=\textsc{ElimConj-L}]
  { \phi \land \psi }
  { \phi }
\quad
\inferrule*[right=\textsc{ElimConj-R}]
  { \phi \land \psi }
  { \psi }
\\[4ex]
\inferrule*[right=\textsc{IntroDisj-L}]
  { \phi }
  { \phi \lor \psi }
\quad
\inferrule*[right=\textsc{IntroDisj-R}]
  { \psi }
  { \phi \lor \psi }
&
\inferrule*[right=\textsc{ElimDisj}]
  { \phi \lor \psi \\ [\phi]^n \;\; \, \cdots \, \;\; \chi \\ [\psi]^m \;\; \, \cdots \, \;\; \chi }
  { \chi }
\\[4ex]
\inferrule*[right=\textsc{IntroImpl}]
  { [\phi]^n \;\; \, \cdots \, \;\; \psi }
  { \phi \to \psi }
&
\inferrule*[right=\textsc{ElimImpl}]
  { \phi \\ \phi \to \psi }
  { \psi }
\\[4ex]
\inferrule*[right=\textsc{IntroNeg}]
  { [\phi]^n \;\; \, \cdots \, \;\; \bot }
  { \neg \phi }
&
\inferrule*[right=\textsc{ElimNeg}]
  { \phi \\ \neg \phi }
  { \bot }
\\[4ex]
&
\inferrule*[right=\textsc{ExFalso}]
  { \bot }
  { \phi }
\end{array}
\]

\section{\centering Alternative Presentation of Natural Deduction}
In the standard presentation of natural deduction, assumptions are introduced as 
leaves of a proof tree and later discharged dynamically. This makes it somewhat 
unclear when and where each assumption is canceled in the finished derivation. 
Texts like \cite{selinger2008:lambdanotes} use an alternative approach to present derivations in terms of judgments,
which are of the form
\[
x_1 : A_1, \dots, x_n : A_n \;\;\vdash\;\; B.
\]
Such a judgment expresses that the formula $B$ can be derived from the 
assumptions $A_1, \dots, A_n$, where each assumption is labeled by a distinct 
variable. Every line in a derivation explicitly carries the full set of 
assumptions used to obtain it. This makes derivations more verbose, but 
assumption management becomes fully transparent.
\subsubsection{\centering New style of Introduction / Elimination Rules}
Let $\Gamma$ denote a context of labeled assumptions. The rules of natural 
deduction can be reformulated as follows:

\[
\begin{array}{@{}l@{\qquad}l@{}}
\inferrule*[right=\textsc{Assume-X}]
  { }
  { \Gamma, x:A \vdash A }
&
\inferrule*[right=\textsc{IntroTop}]
  { }
  { \Gamma \vdash \top }
\\[4ex]
\inferrule*[right=\textsc{ElimConj-L}]
  { \Gamma \vdash A \land B }
  { \Gamma \vdash A }
\quad
\inferrule*[right=\textsc{ElimConj-R}]
  { \Gamma \vdash A \land B }
  { \Gamma \vdash B }
&
\inferrule*[right=\textsc{IntroConj}]
  { \Gamma \vdash A \\ \Gamma \vdash B }
  { \Gamma \vdash A \land B }
\\[4ex]
\inferrule*[right=\textsc{IntroDisj-L}]
  { \Gamma \vdash A }
  { \Gamma \vdash A \lor B }
\quad
\inferrule*[right=\textsc{IntroDisj-R}]
  { \Gamma \vdash B }
  { \Gamma \vdash A \lor B }
&
\inferrule*[right=\textsc{ElimDisj}]
  { \Gamma \vdash A \lor B \\ \Gamma, x:A \vdash \chi \\ \Gamma, y:B \vdash \chi }
  { \Gamma \vdash \chi }
\\[4ex]
\inferrule*[right=\textsc{IntroImpl}]
  { \Gamma, x:A \vdash B }
  { \Gamma \vdash A \to B }
&
\inferrule*[right=\textsc{ElimImpl}]
  { \Gamma \vdash A \to B \\ \Gamma \vdash A }
  { \Gamma \vdash B }
\\[4ex]
\inferrule*[right=\textsc{IntroNeg}]
  { \Gamma, x:A \vdash \bot }
  { \Gamma \vdash \neg A }
&
\inferrule*[right=\textsc{ElimNeg}]
  { \Gamma \vdash A \\ \Gamma \vdash \neg A }
  { \Gamma \vdash \bot }
\\[4ex]
&
\inferrule*[right=\textsc{ExFalso}]
  { \Gamma \vdash \bot }
  { \Gamma \vdash \chi }
\end{array}%
\]
\begin{example} Comparing the two alternative styles
  \label{ex:traditionalvsalternative}
\[
\begin{array}{@{}c@{\qquad}c@{}}
\text{Traditional} &
\text{Explicit Contexts}
\\[2ex]
\inferrule*[right=\textsc{IntroImpl}]
  { \inferrule*[right=\textsc{ElimConj-L}]
      { [A \land B] }
      { A }
  }
  { \vdash (A \land B) \to A }
&
\inferrule*[right=\textsc{IntroImpl}]
  { \inferrule*[right=\textsc{ElimConj-L}]
      { \inferrule*[right=\textsc{Ax}]
          { }
          { x:A \land B \;\vdash\; A \land B }
      }
      { x:A \land B \;\vdash\; A }
  }
  { \vdash (A \land B) \to A }
\end{array}
\]
\end{example}

\begin{note} Going over Example~\ref{ex:traditionalvsalternative}, we observe:
\begin{itemize}
  \item In the traditional style, we assume $A \land B$ at the leaf, apply \textsc{ElimConj-L} to obtain $A$, and then discharge the assumption with \textsc{IntroImpl}.
  \item In the explicit-context style, each judgment carries its assumptions. The axiom \textsc{Assume-X} gives $x:A \land B \vdash A \land B$, then \textsc{ElimConj-L} gives $x:A \land B \vdash A$, and finally \textsc{IntroImpl} discharges the labeled assumption $x:A \land B$.
\end{itemize}
\end{note}
  
\begin{note}
In the traditional style, assumptions appear only at the leaves and are canceled 
dynamically. In the alternative style, contexts are explicitly tracked in every 
judgment, making assumption management fully transparent at the cost of greater 
verbosity.
\end{note}
\section{\centering The Curry--Howard Isomorphism}
The $\lambda$-calculus, as introduced in the first section, provides a minimal but expressive framework for modeling computation. Independently, intuitionistic logic offers a formal system for constructive reasoning that requires the existence of witnesses for propositions. The Curry-Howard correspondence reveals a structural equivalence between them, providing a unifying perspective that connects proofs with programs. The Curry--Howard correspondence establishes the following associations:
\begin{itemize}
\item \textbf{Propositions $\leftrightarrow$ Types}: Proposition in intuitionistic logic can be interpreted as a type in a typed $\lambda$-calculus.
\item \textbf{Proofs $\leftrightarrow$ Programs}: Constructive proofs of a proposition correspond to a $\lambda$-term inhabiting the associated type.
\item \textbf{Proof normalization $\leftrightarrow$ Program evaluation}: The process of simplifying proofs  corresponds to the process of evaluating a term through $\beta$-reduction.
\end{itemize}
\subsubsection{\centering \textit{``Proofs as Programs''}}
Given a proposition $P$ in intuitionistic logic, constructing a proof of $P$ is equivalent to constructing a term $t$ such that $t : T$, where $T$ is the type corresponding to $P$. This yields the principle often summarized as \emph{``proofs-as-programs''} or \emph{``propositions-as-types''}. The specifics of this relation are summarized in table \ref{tab:curry-howard}.

\begin{table}[h!]
  \centering
    \caption{The Curry--Howard Correspondence}
  \begin{tabular}{|c|c|}
    \hline
    \textbf{Intuitionistic Logic} & \textbf{Typed $\lambda$-Calculus} \\
    \hline
    Proposition $A$ & Type $A$ \\
    Proof of $A$ & Term of type $A$ \\
    Implication $A \rightarrow B$ & Function type $A \to B$ \\
    Conjunction $A \land B$ & Product type $A \times B$ \\
    Disjunction $A \lor B$ & Sum type $A + B$ \\
    $\top$ & $1$ \\
    $\bot$ & $0$ \\
    Proof normalization & $\beta$-reduction \\
    \hline
  \end{tabular}
  \label{tab:curry-howard}
\end{table}

\begin{example}
Consider the implication $A \rightarrow B$ in intuitionistic logic. A proof of $A \rightarrow B$ is a construction that, given a proof of $A$, produces a proof of $B$. Under the Curry--Howard correspondence, this directly maps to a function type $A \to B$ in $\lambda$-calculus, where a $\lambda$-term $f : A \to B$ takes an argument of type $A$ and returns a result of type $B$. Similarly, conjunction $A \land B$ corresponds to a product type $A \times B$, and disjunction $A \lor B$ corresponds to a sum type $A + B$. In each case, the rules for constructing and deconstructing proofs mirror the operations on the corresponding $\lambda$-terms.
\end{example}
\begin{example} Modus Ponens via Curry--Howard. The logical rule for modus ponens is:
\[
\inferrule*[right=\textsc{MP}]
  {P \to Q \quad P}
  {Q}
\]
Walking our way back through table \ref{tab:curry-howard}, the $\la$--term should be of type $(P \to Q) \to P \to Q$, since it takes a proof of $P \to Q$ and a proof of $P$ and produces a proof of $Q$. Also, we can inhabitate the type using the fact that applying a term of type $(P \to Q)$ to a term of type $P$ yields a term of type $Q$ by \textsc{App} in the typed \lcalc. The rule \textsc{MP} corresponds to the following typed term:
\[
  \vdash \la f.\, \la x.\, f\, x : (P \to Q) \to P \to Q
\]
Step-by-step $\beta$-reduction:
\[
(\la f.\, \la x.\, f\, x)\ f\ x
\;\;\curly\;\; (\la x.\, f\, x)\ x
\;\;\curly\;\; f\, x
\]
This reduction might not say much by itself, so let us annotate it with the most general type of each of the terms in the reduction:
\[
\begin{aligned}
&(\lambda f : P \to Q.\, \lambda x : P.\, f\, x)\;\; f : P \to Q \;\; x : P 
&&\vdash (P \to Q) \to P \to Q \\
&\;\;\curly\;\;
(\lambda x : P.\, f\, x) \;\; x : P 
&&\vdash P \to Q \\
&\;\;\curly\;\;
f\, x 
&&\vdash Q 
\end{aligned}
\]
\end{example}
\paragraph{References \& Resources.}{
  An accessible entry point to intuitionistic logic and natural deduction is given in the lecture notes by \cite{selinger2008:lambdanotes}, it is rigorous yet pedagogical. A more modular and systematically updated resource is the Open Logic Project~\cite{openlogicprojectintuitionisticlogic2025}.
  Girard's \emph{Proofs and Types}~\cite{girard1989:proofs-and-types} has also influenced this chapter, as well as Sørensen and Urzyczyn’s \emph{Lectures on the Curry--Howard Isomorphism}~\cite{sorensen2006:curry-howard}, that presents the foundational connections between intuitionistic logic, proof theory, and type theory.

