

\chapter*{Abstract}


\noindent This work investigates the interplay between $\lambda$-calculus, type theory, and category theory. It is written for readers without a deep mathematical background who nevertheless seek a rigorous but accessible introduction. The exposition emphasizes $\lambda$-calculus both as a model of computation and as the foundation for modern type systems, discussing essential properties such as the Church--Rosser theorem and normalization.

Typing disciplines are explored in detail, showing how types act as logical propositions and programs as their proofs, in accordance with the Curry--Howard correspondence. Type theory is then connected to category theory via the structure of cartesian closed categories, showing how categorical semantics provide a unifying perspective.

Throughout, attention is given to why the $\lambda$-calculus is a suitable model of computation, how its connection to types and categories enriches our understanding, and how these ideas influence the design of programming languages and related areas, in line with the broader Curry--Howard--Lambek correspondence.

\vspace*{0.7cm}

\noindent
\textit{Este trabajo investiga la relación entre el cálculo} $\lambda$\textit{, la teoría de tipos y la teoría de categorías. 
Está escrito para lectores sin una formación matemática profunda, pero que aun así buscan una introducción rigurosa y accesible. 
La exposición pone énfasis en el cálculo} $\lambda$ \textit{tanto como modelo de computación como fundamento de los sistemas de tipos modernos, discutiendo propiedades esenciales como el teorema de Church--Rosser y la normalización.}

\textit{Las disciplinas de tipado se exploran en detalle, mostrando cómo los tipos actúan como proposiciones lógicas y los programas como sus pruebas, de acuerdo con la correspondencia de Curry--Howard. 
Posteriormente, la teoría de tipos se conecta con la teoría de categorías a través de la estructura de las categorías cartesianas cerradas, mostrando cómo la semántica categórica ofrece una perspectiva unificadora.}

\textit{A lo largo del texto se presta atención a por qué el cálculo} $\lambda$ \textit{es un modelo adecuado de computación, cómo su vínculo con los tipos y las categorías enriquece nuestra comprensión, y de qué manera estas ideas influyen en el diseño de los lenguajes de programación y áreas relacionadas, en el marco de la más amplia correspondencia de Curry--Howard--Lambek.}

\vspace*{1cm}

\noindent \textbf{Keywords}:
Lambda Calculus,
Type Theory,
Category Theory,
Intuitionistic Logic,
Curry-Howard-Lambek Correspondence,
Closed Cartesian Categories.

\newpage

