\chapter*{Introduction}

This work means to provide an introduction to functional programming and categorical reasoning about programs, using the \lcalc \ as its foundational formalism for the study of computation. With this in mind, we cover confluence and normalization properties to an acceptable level of detail, as to provide the reader with enough evidence that this system is solid enough.

In its untyped form, the \lcalc \ lacks the structure necessary to guarantee properties such as safeness and termination. This lack of structure, naturally leads into typing. We cover monomorphic type systems and the notion of \textit{``programs as proofs''} as introduced by \cite{Howard1980}. Typing disciplines enrich the \lcalc\ by imposing constraints on expressions, ensuring that only well-formed programs can be constructed, guaranteeing termination and safety. Type systems also give compositional structure to programs, such that given
\(
g : A \to B, f : B \to C
\) we can construct \(
f \circ g : A \to C.
\)  
The meaning of this composition is entirely determined by the meanings of \(f\) and \(g\) and the rule of composition. Compositionality shifts the focus from specific objects to the morphisms between them, and is arguably the core idea behind functional programming.

This principle---where the whole is defined by its parts and their composition---extends beyond functions between sets. Category theory abstracts this idea: it generalizes functions to morphisms and provides a unified language for compositional structure. The work by \cite{Lambek1980} motivates the final chapters of this work, presenting Closed Cartesian Categories, and how they extend the Curry--Howard Isomorphism, giving a categorical interpretation of the work by Curry and Howard.

The following is informal but approachable summary to this work.
This summary is intended as an overview of the landscape connecting type theory 
and category theory. This is because
while we cover the core computational properties of the $\lambda$-calculus, the final goal of the document
is to make some higher-level connections. To this end, this work emphasizes computation first, 
and then gradually abstracts toward categorical perspectives as the exposition 
develops. To do justice to the title, the summary presents the
higher-level abstractions that connect the $\lambda$-calculus to type theory and category 
theory. So that the reader understands the discussion going on behind the concrete notions
of computation and have an overview of the increasingly abstract perspectives.

\begin{comment}
\orange{
\input{sections/introduction-summary}
}
\end{comment}

\subsubsection*{\centering Summary}

In order to get acquainted with the \lcalc, let us develop a simple example to familiarize ourselves before we begin with a more formal approach to this discipline. Consider the function $f(x) = x + 1$, the most straightforward way to express this using \lcalc \ would be $(\la x . x + 1 )$, where the lambda denotes that $x$ is being captured and used as a parameter to perform some computation. Evaluating $f(2)$ using this newly created lambda term would look like this: $(\la x . x + 1)(2) \mathrel{\leadsto} (2 + 1) \mathrel{\leadsto} 3$. This process of reducing a $\la$-expression is referred to as \bred.

The untyped $\lambda$-calculus is extremely expressive but gives no static guarantees about program behaviour --- in particular, divergent terms exist, like the $\Omega$ combinator \ref{ex:omega}. To regain discipline we annotate terms with types and move to the simply typed $\lambda$-calculus. We start from a collection of basic types $A,B,C,\dots$. From these we form arrow types:
\[
  \text{if }A\text{ and }B\text{ are types, then }A \to B \text{ is a type}
\]
Intuitively, a term of type $A \to B$ is a function which, given an input of type $A$, returns an output of type $B$. Because reduction is only performed on well-typed applications, types enforce that compositions are safe: if $M : A \to B$ and $N : A$, then $M\,N$ has type $B$. For example,
\[
  \lambda x.\,x+1 : \mathbb{N} \to \mathbb{N}
  \qquad
  (\lambda x.\,x+1)\;2 \;\leadsto\; 3
\]
so the domain is $\mathbb{N}$ and the result is a $\mathbb{N}$. In short: annotating terms with arrow types restricts how functions can be combined and yields a safer, more predictable calculus. This idea of types can be further extended. The extensions we cover in depth are:

\begin{itemize}
\item A product type $A \times B$ in the simply typed $\lambda$-calculus means ``having both an $A$ and a $B$ at the same time''. If $a : A$ and $b : B$, then $\langle a,b\rangle : A \times B$. A product is equipped with projections $\pi_{1}\langle a,b \rangle = a$, $\pi_{2}\langle a,b\rangle = b$. In Haskell:
\begin{minted}{Haskell}
  data Pair a b = Pair a b -- Product Type: A x B
  
  pi1 :: Pair a b -> a -- Projections
  pi1 (Pair x _) = x
  pi2 :: Pair a b -> b
  pi2 (Pair _ y) = y
\end{minted}

Intuitively, this is just like a tuple where $\pi_{1}$ picks the first component and $\pi_{2}$ picks the second.

\item In contrast, a sum type $A + B$ in the simply typed $\lambda$-calculus means ``having either an $A$ or a $B$, and knowing which one it is''.
Introduction: if $a : A$ then $\iota_{1} a : A + B$; if $b : B$ then $\iota_{2} b : A + B$. A case analysis that handles both possibilities:  
\[
  \text{case } s \text{ of } 
    \iota_{1}(a) \mapsto M \;\;|\;\;
    \iota_{2}(b) \mapsto N
  \]
  
Intuitively, this is a tagged union: a value carries a label and a payload. In Haskell:
\begin{minted}{Haskell}
  data Sum a b = InL a | InR b -- Sum type: A + B

  inl :: a -> Sum a b -- Injections
  inl = InL
  inr :: b -> Sum a b
  inr = InR
\end{minted}

\end{itemize}

This extensions allow similar behavior to algebraic types in Haskell, the key difference is that Haskell also allows type definitions, where a type can be defined in terms of itself. The simply typed $\lambda$-calculus with sums and products does not have this. For example, lists in Haskell are recursive:  
\begin{minted}{haskell}
  data List a = Nil | Cons a (List a)
\end{minted}

This says a list is either empty or a pair of an element and another list.  
Such recursive types go beyond what can be expressed with just sums and products in the simply typed $\lambda$-calculus and what we will cover.

\medskip

The simply typed $\lambda$-calculus with products and sums has a direct logical correspondence where types correspond to propositions, and terms correspond to proofs. This interpretation is called the Curry--Howard isomorphism. The product type $A \times B$ behaves like logical conjunction $A \land B$, and the sum type $A + B$ behaves like logical disjunction $A \lor B$. Establishing that programming with functions and data is formally the same as reasoning in propositional logic.

\medskip

Up until now we have summarized the role behind the type theoretic grounds that this work covers and how it connects to logic, but this work is intended as a ``Unified Approach'' to type and category theory. We follow up by informally presenting what categories are and some high level notions of how they relate to types.

\medskip

Category theory studies objects indirectly, rather than probing the internal nature of an object, one examines how it relates to others, what arrows it admits, and how these compose. In spirit this is not far from the proverb \emph{``dime de quiénes te rodeas y te diré quién eres''}.

Just as types discipline how terms can be combined in the simply typed $\lambda$-calculus, 
categories discipline how objects can be related through arrows also called morphisms.
A category consists of: a collection of objects, a collection of arrows or morphisms between objects and a few important rules about composition. A familiar example is the category \textbf{Set}, where objects are sets, arrows are total functions between sets, composition is ordinary function composition.

To begin shifting the readers focus from objects to morphisms, it is helpful to see how familiar notions like injectivity can be re-expressed without referring to elements at all. In the category \textbf{Set}, one usually defines a function as injective by saying that distinct elements of the domain are always mapped to distinct elements of the codomain. Category theory instead abstracts this idea: an arrow $f : A \to B$ is called a monomorphism if for any pair of arrows
$g_1, g_2 : C \to A$
\[
   f \circ g_1 = f \circ g_2 \quad \Longrightarrow \quad g_1 = g_2
 \]
 
In other words, once composed with $f$, different arrows into $A$ remain distinguishable. 
This captures exactly the behavior of injective functions in \textbf{Set}, 
but it does so purely in terms of morphisms and composition. 
The advantage of this perspective is that it applies uniformly to any category, 
even in contexts where it makes no sense to speak about individual elements of objects.

In the same way that injectivity can be reformulated in terms of morphisms, 
constructions like pairs and disjoint unions are captured categorically 
by the notions of product and coproduct, which embody the behavior of 
tuples and sums purely through their universal properties of arrows. Beyond sums and products, categories also support exponentials, which capture the idea of 
function spaces. Intuitively, an exponential $B^A$ represents ``the object of arrows'' from $A$ to $B$. 
It comes equipped with an evaluation morphism
\[
   \mathsf{eval} : B^A \times A \to B,
\]
expressing the idea that a pair of a function and an argument produces a result. 
This abstractly mirrors how function application works in the $\lambda$-calculus.

Putting these ideas together, we arrive at the notion of a 
Cartesian Closed Category: a category equipped with products, a terminal object, 
and exponentials. Such categories provide the natural semantics of the simply typed 
$\lambda$-calculus, where types correspond to objects, terms correspond to morphisms, 
and function application is captured by the evaluation morphism.

This ongoing interaction between type theory and category theory has inspired the design 
of entirely new calculi and semantic frameworks. A notable example is the work of 
Cockett and Gallagher (2019), who developed \emph{Categorical models of the differential 
$\lambda$-calculus} within the setting of differential categories. Their results show how 
categorical structure can guide the extension of the $\lambda$-calculus with new operations 
such as differentiation, providing both a sound semantics and fresh perspectives on 
computation. This illustrates how category theory is not merely a language for 
describing existing systems, but also a tool for discovering and shaping new ones.

Today, the influence of categorical semantics on programming languages is unmistakable. 
Functional languages such as Haskell, OCaml, and Coq are designed on principles that arise 
directly from type theory and categorical reasoning, including higher-order functions, 
algebraic data types, and polymorphism. Monads, for instance, originated as a categorical 
concept and have become central for structuring effects in modern programming. 
More recent advances such as dependent types, homotopy type theory, and effect systems 
show that the dialogue between category theory and type theory continues to inform both 
the foundations of logic and the design of real programming languages. 
This ongoing exchange ensures that the state of the art in programming is firmly connected 
to deep mathematical ideas, with categorical semantics providing both the language and 
the tools to innovate further.

\paragraph{Contents.}
{
As already discussed, this work explores the untyped and typed lambda calculus and the Curry--Howard Isomorphism. We then expand the correspondence by adding categorical semantics to build what is referred to as the Curry--Howard--Lambek Correspondence.
}

\begin{itemize}
\item The first chapter is an introduction to the \lcalc. It emphasizes confluence as a pillar for computation, it also introduces normalization as a desirable property.

\item In the second chapter, typing is introduced, together with its so desired normalization properties. Also, differences with the untyped \lcalc \ are discussed.

\item The third chapter glances over intuitionistic logic and natural deduction. It introduces the Curry--Howard correspondence.

\item The fifth chapter is about category theory and its constructions. We cover functors, natural transformations and adjunctions. We also cover the necessary universal constructions to build Closed Cartesian Categories, namely, terminal objects, products and exponentials.

\item The last chapter extends the Curry--Howard correspondence, establishing a relation between \lcalc \ and Closed Cartesian Categories.
\end{itemize}


\paragraph{Target Audience.}{
This text is intended for computer scientists who may be unfamiliar with category theory but are interested in understanding the theoretical foundations of functional programming. It aims to provide readers with enough background on typing so that they can better understand the relation functional programming languages have with category theory, compositionality, type systems and other concepts inside of Category Theory that drive the development of sate-of-the-art programming.

This text assumes basic knowledge of mathematics, familiarity with programming, and a working understanding of recursion. No prior experience with functional programming or category theory is required, though some familiarity would not hurt.
}

\paragraph{Rationale behind our Approach.}{
  This text occupies a middle ground between works like \cite{Milewski2019} and \cite{maclane1971categories}. The former emphasizes intuitiveness, and while it serves as a powerful toolkit for programmers, it gives little insight into the formalisms of categorical reasoning. The latter, presents category theory in full generality, making it inaccessible to non-mathematicians. Our approach introduces the fundamental notions in a somewhat mathematically precise manner first to then develop programmer-friendly examples, applications, and intuitions. This text is of course not in the same league as the aforementioned, it is a short and concise introduction and should be read as such.
}

\paragraph{External Resources.}
{
The code is meant as a learning aid, nevertheless, it can be given purpose. Similarity to the formal definitions is the driving design principle behind most of the code, as well as ease of use. We use the various programs to simplify some of the examples and avoid including unnecessarily large reduction procedures.
\begin{itemize}
  
\item To clone the \LaTeX \ source used to generate this document:
\begin{minted}{shell-session}
  $ git clone https://github.com/laredo02/ttcmua.git
\end{minted}

\item To clone the code used to implement the \lcalc:
\begin{minted}{shell-session}
  $ git clone https://github.com/laredo02/lcalc.git
\end{minted}

There are a few different files in the directory, each of which encapsulates a different approach to typing:
\begin{itemize}
\item Untyped
\item Fully annotated typing
\item Typed with annotations only on abstractions i.e. Typing à la Church
\item Typed with monomorphic type inference i.e. Typing à la Curry
\end{itemize}
Files with miscellaneous examples related to categories are also included, in particular those about functors, natural transformations and their composition may be of interest. Implementations are supposed to especially help programmers to digest some of the contents of this work. To use the provided tools, we recommend running them on \texttt{ghci}, Haskell's interactive shell.


\end{itemize}
}

\paragraph{Disclosure.}
{
  In the development of this work, I have employed AI tools to assist me with typesetting and at times to look for synonyms, as English is not my native tongue. All the content, including explanations, examples, and discussions, is original and developed by me. Whenever external material has been incorporated, it has been cited appropriately. At all times, I have verified the correctness of the information up to my level of competence, ensuring that the final work reflects my own understanding and effort, which is the ultimate goal of the document.
}



