\section{\centering Encodings and Constructions}

This section presents a number of conventional constructions that represent key ideas common to computer science and other fields. We are specifically referring to natural numbers, branching statements, booleans and lists among others, as well as the operations over them. These constructions can be used to prove relevant statements concerning the expressiveness of the $\la$-calculus.

\begin{note}
  We use our Haskell program to bring some computer science constructions to life, which means that some of the details in this section are implementation dependent. In addition, we use the following type aliases to simplify notation: \texttt{v} \; \texttt{=} \; \texttt{Var;} \texttt{l} \; \texttt{=} \; \texttt{Abs;} \texttt{a} \; \texttt{=} \; \texttt{App}.
\end{note}

\subsubsection{\centering Natural Numbers}
We briefly recall the axiomatic characterization of the natural numbers due to Peano. The Peano axioms provide a minimal foundation for arithmetic, based on a constant $0$, a successor function $S$, and induction.

The connection with the $\lambda$-calculus arises from the fact that the same structure can be represented purely functionally. By encoding $0$, successor, we recover the essential properties of natural numbers without referring to any external notion of arithmetic. This construction, known as Church encoding, illustrates the expressive power of the $\lambda$-calculus and provides the groundwork for representing data and computation purely in terms of functions.

\begin{definition} Peano Axioms, the set of natural numbers $\mathbb{N}$
\begin{itemize}
\item \textbf{(Zero)} \( 0 \in \NAT \)
\item \textbf{(Successor)} \( \forall n \in \NAT, S(n) \in \NAT \)
\item (Initial) \( \forall n \in \NAT, S(n) \neq 0 \) \label{def:peano-nat-initial}
\item (Injectivity) \( \forall n, m \in \NAT, S(n) = S(m) \Rightarrow n = m \)
\item (Induction) \( P(0) \land \forall ( P(n) \Rightarrow P(S(n))) \Rightarrow \forall n,  P (n) \)
\end{itemize}
The first four axioms can be summarized using a graph:
\end{definition}
\begin{example}
  Say for instance we mean to represent $ 3 $ using Peano notation, using the graph, we would come up with $ s(s(s(0))) $, since it is $3$ steps away from $0$.
\end{example}
\begin{definition} To add and multiply natural numbers we use following the rules:
  \[
    n + m = 
    \begin{cases}
      n & \text{if } m = 0 \\
      s(n + m') & \text{if } m = s(m')
    \end{cases}
    \quad
    a \cdot b =
    \begin{cases}
      0 & \text{if } b = 0 \\
      a \cdot b' + a & \text{if } b = s(b')
    \end{cases}
  \]
\end{definition}
\begin{definition} Representation of the set of all naturals in the \lcalc, usually referred to as Church numerals:
  \label{def:church-naturals}
\[
    \overline{\texttt{0}} \equiv \la f.\la x. x \qquad \texttt{SUCC} \equiv \la n. \la f. \la x . f ( n f x )
\]
  Using our Haskell program, $\overline{\texttt{0}}$ is represented like follows:
  \begin{minted}{Haskell}
  zero  = l "f" (l "x" (v "x"))
  \end{minted}
\end{definition}
\begin{example}
  The church numerals have been defined similarly to the peano axioms, that is, by bringing into existance an initial element and using the succesor function to define the rest of them. Let us apply \texttt{SUCC} a few times to illustrate:
\[
    \overline{\texttt{0}} \equiv \la f.\la x. x \qquad
    \overline{\texttt{1}} \equiv \la f.\la x. f x \qquad
    \overline{\texttt{2}} \equiv \la f.\la x. f (f x) \qquad
    \overline{\texttt{n}} \equiv \la f.\la x. f ^ n x
\]
  where $ n $ denotes the number of times we append $ f $.
\end{example}
\begin{notation}
  We use $\overline{\text{number}}$ to denote a number is church encoded. In our Haskell program we use \texttt{nat} to convert from Haskell's prelude type \texttt{Natural} to type our custom type \texttt{Term}.
\begin{minted}{haskell}
  nat :: Natural :: Term
  nat n = normalize (nat_ n)
      where
          nat_ 0 = zero
          nat_ n = a succ_ (nat_ (n - 1))
\end{minted}
\end{notation}

\begin{remark}
  Since \lterms \ are first-class citizens, and $f$ and $x$ are abstracted away, the reality is that numbers are a quite general construct that has many distinct applications. Recall Example \ref{ex:high-order-by-example} where $A \equiv_\alpha \overline{\texttt{2}}$.
\end{remark}
\subsubsection{\centering Basic Arithmetic}
\begin{definition} Now that we have a construction for the naturals, we can define some arithmetic operations:
\[
  \texttt{ADD} \equiv \la m.\la n. \la f. \la x. m f ( n f x) \qquad
  \texttt{MUL} \equiv \la f.\la x. \la f. \la x. m (n f) x
  \]
  \[
    \texttt{ISZERO} \equiv \lambda n. n\,(\lambda x.\texttt{FALSE})\,\texttt{TRUE} \qquad
    \texttt{PRED} \equiv \la n.\,\la f.\,\la x.\, \bigl(((n \; (\la g.\,\la h.\, h\,(g\,f))) \; (\la u.\,x)) \;
    (\la u.\,u))\bigr)
  \]
\texttt{PRED} computes the predecessor of a church encoded natural. $\texttt{PRED} \, \overline{0} = \overline{0}$. We omit further Haskell implementation of constructions from now on since they are just literal translations to our programs representation of \lterms.
\end{definition}
\begin{example} Now we illustrate how addition works by performing $2 + 3$ with the system we defined.
    \[
    \begin{aligned}
      \texttt{ADD}\;\texttt{2}\;\texttt{3} &\equiv (\la m.\la n.\la f.\la x. m\,f\,(n\,f\,x))\;(\la f.\la x. f\,f\,x)\;(\la f.\la x. f\,f\,f\,x) \\
                                           &\curly^* \la f.\la x.\; (\la f.\la x. f\,f\,x)\;f\;((\la f.\la x. f\,f\,f\,x)\;f\;x) \\
                                           &\curly^* \la f.\la x.\; (\la x. f\,f\,x)\;((\la x. f\,f\,f\,x)\;x) \\
                                           &\curly (\la f.\la x.\; f\,f\,f\,f\,f\,x) \\
                                           & \equiv \texttt{5}
    \end{aligned}
  \]
  In Haskell:
  \begin{minted}{shell-session}
  ghci> normalize (a (a mul (nat 2)) (nat 3))
\end{minted}
This unequivocally yields \texttt{(\lambda f.(\lambda x.f(f(f(f(f(fx)))))))}.
\end{example}
\begin{remark}
  Multiplication works similarly to addition. As for \texttt{ISZERO}, the intuition behind it is that if the number is zero, then \(\lambda x.\, \texttt{FALSE}\) is dropped, and the result is \texttt{TRUE}; otherwise, it is applied at least once, which immediately yields \texttt{FALSE}.
  \end{remark}
\begin{remark}
  The \lcalc \ is at least capable of arithmetic.
\end{remark}
\subsubsection{\centering Booleans \& Branching}
The next constructions in line are the truth values of boolean logic, ``\textit{true}'' and ``\textit{false}'' for short.
\begin{definition} Boolean truth values:
\(
    \texttt{TRUE} \equiv \lambda t.\lambda f. t \text{ and }
    \texttt{FALSE} \equiv \lambda t.\lambda f. f
\)
\end{definition}
\begin{remark}
  \label{rem:true-false}
  Note how $\TRUE \ M \ N \curly M $ and $ \FALSE \ M \ N \curly N $
\end{remark}
\begin{definition} Conditional, in other words \textit{``if statement''}.
  \(
    \texttt{IF} \equiv \lambda b.\lambda t.\lambda e. b\,t\,e \)
\end{definition}
\begin{note}
  \IF \ statement is just a combinator that happens to behave similarly to a branching statement whenever we pass a church encoded boolean as a first argument.
  \end{note}
\begin{example} On the correct behabiour of $\IF$ i.e. check that it implements branching. $ M, N \in \La $ and $ \IF, \TRUE, \FALSE$ are as usual:
  \[
    \begin{aligned}
      & \IF \ \TRUE \ M \ N \\
      \equiv \ & (\la b.\la t.\la e.\, b\, t\, e) \ \TRUE \ M \ N  \\
      \curly \ & (\la t.\la e.\, \TRUE\ t\, e) \ M \ N  \\
      \curly \ & (\la e.\, \TRUE\ M\, e) \ N  \\
      \curly \ & \TRUE\ M\, N \curly^* M 
    \end{aligned}
    \quad
    \begin{aligned}
      & \IF \ \FALSE \ M \ N \\
      \curly^* \ & \FALSE\ M\, N \curly^* N
    \end{aligned}
  \]
  So the \IF \ just combines $b, t, e$ in such a way that whenever b is a church encoded boolean either $t$ or $e$ is dropped depending on the truth value of $b$. See Remark \ref{rem:true-false}. From this example we extract that $ \la x . x $ is equivalent to \IF \ it terms of behaviour, and thus, we could use $ \IF \equiv \la x . x $ interchangeably.
\end{example}
\begin{definition}
  Using the \IF \ we can implement an universal set of logic gates:
\[
    \AND \equiv \IF \, a \, (\IF \, b \, \TRUE \, \FALSE) \, \FALSE \qquad
    \OR \equiv \IF \, a \, \TRUE \, \TRUE \, (\IF \, b \, \TRUE \, \FALSE) \qquad
    \NOT \equiv \IF \, b \, \FALSE \, \TRUE
\]
This allows us to simulate a nand gate, known to be universal. Thus, the lambda calculus is at least equivalent to propositional logic.
\end{definition}

\begin{definition} Define the translation $\llbracket\cdot\rrbracket$ that translates a formulas from propositional logic to their Church encoded representation:
  \label{def:proptocalc}
  \[
    \begin{aligned}
      \llbracket p_i\rrbracket &= P_i \quad(\text{a }\lambda\text{-variable})\\
      \llbracket \neg\varphi\rrbracket &= \NOT\,\llbracket\varphi\rrbracket\\
      \llbracket \varphi\wedge\psi\rrbracket &= \AND\,\llbracket\varphi\rrbracket\,\llbracket\psi\rrbracket\\
      \llbracket \varphi\vee\psi\rrbracket &= \OR\,\llbracket\varphi\rrbracket\,\llbracket\psi\rrbracket
    \end{aligned}
  \]
\end{definition}

\begin{proposition}
  For every propositional formula $\varphi$, in particular every conjunctive normal form,
  there is a $\lambda$-term $\llbracket\varphi\rrbracket$ built from the
  variables and the terms $\AND,\OR,\NOT$ such
  that for each Boolean assignment $\sigma$, substituting each propositional
  variable by $\TRUE$ or $\FALSE$ yields:
  \[
    \varphi(\sigma)=\mathsf{true}\iff \llbracket\varphi\rrbracket[\sigma]\curly^* \TRUE
    \qquad
    \varphi(\sigma)=\mathsf{false}\iff \llbracket\varphi\rrbracket[\sigma]\curly^* \FALSE
  \]
  Thus the $\lambda$-calculus with these terms can simulate any conjunctive normal form and
  hence propositional logic.
\end{proposition}

\begin{proof} Any conjunctive normal form is a formula built from $\neg,\vee,\wedge$, so we can use the representation \ref{def:proptocalc} that translates from propositional logic in conjunctive normal form into Church encoded terms. We aim to prove the following invariant by structural induction on $\varphi$:
  \[
    \llbracket\varphi\rrbracket[\sigma]\curly^*\TRUE
    \iff \varphi(\sigma)=\mathsf{true}
  \]
  and the analogous statement for $\FALSE$. Where $\sigma$ is a Boolean assignment and $\llbracket\varphi\rrbracket[\sigma]$ stands for the substitution that replaces each $P_i$ by $\sigma(p_i)\in\{\TRUE,\FALSE\}$.

  \begin{itemize}

  \item Base case: If $\varphi=p_i$ then $\llbracket p_i\rrbracket[\sigma]=\sigma(p_i)$,
    so the claim holds by definition of $\sigma$.

  \item Inductive step: Assume the claim for $\varphi$ and $\psi$ for the inductive hypothesis.
    \begin{itemize}
    \item For negations we have $\llbracket\neg\varphi\rrbracket[\sigma]=\NOT\,\llbracket\varphi\rrbracket[\sigma]$.
      By the inductive hypothesis $\llbracket\varphi\rrbracket[\sigma]\curly^*\TRUE$
      if and only if $\varphi(\sigma)=\mathsf{true}$. Since $\NOT$ inverts Church Booleans
      i.e. $\NOT\,\TRUE\curly^*\FALSE$ and
      $\NOT\,\FALSE\curly^*\TRUE$, it follows that
      $\llbracket\neg\varphi\rrbracket[\sigma]\curly^*\TRUE$ if and only if
      $\varphi(\sigma)=\mathsf{false}$, i.e. $\neg\varphi(\sigma)=\mathsf{true}$.
    \item For conjunction, $\llbracket\varphi\wedge\psi\rrbracket[\sigma]
      =\AND\,\llbracket\varphi\rrbracket[\sigma]\,\llbracket\psi\rrbracket[\sigma]$.
      By the inductive hypothesis the two subterms reduce to the Church booleans corresponding to
      $\varphi(\sigma)$ and $\psi(\sigma)$. The definition of $\AND$
      ensures it reduces to $\TRUE$ exactly when both inputs are
      $\TRUE$, and to $\FALSE$ otherwise. Hence if and only if $\varphi(\sigma)=\psi(\sigma)=\mathsf{true}$,
      i.e. $(\varphi\wedge\psi)(\sigma)=\mathsf{true}$.
    \item We apply similar reasoning on disjunctions using $\OR$ shows
      $\llbracket\varphi\vee\psi\rrbracket[\sigma]\curly^*\TRUE$
      iff $\varphi(\sigma)=\mathsf{true}$ or $\psi(\sigma)=\mathsf{true}$.
    \end{itemize}
  \end{itemize}
  Thus the invariant holds for all formulas. In particular it holds for conjunctive normal forms,
  so the translated $\lambda$-term evaluates to $\TRUE$ exactly when
  the conjunctive normal form is true under the given assignment. This completes the proof.
\end{proof}

\begin{corollary}
  Sice the \lcalc \ can simulate any conjunctive normal form it is equivalent to propositional logic.
\end{corollary}

\subsubsection{\centering Pairs}

A fundamental construct for building data structures is the notion of pairs. 
They allow us to bind two values together and query them independently.

\begin{definition}
  \( \PAIR \equiv \la x.\la y.\la f. f\,x\,y \).
\end{definition}

\begin{remark}
  This definition is convenient since 
  \(
    \PAIR \, M \, N \curly^* \la f . f \, M \, N
  \).
  Now, we can retrieve the first component via 
  \( (\la f . f \, M \, N) \, \TRUE \) 
  and the second via 
  \( (\la f . f \, M \, N) \, \FALSE \). 
  This motivates the following projection operators.
\end{remark}

\begin{definition}
  A \PAIR \ is nothing but an ordered relation $\langle A, B \rangle \in A \times B$.  
  The projections $\pi_1, \pi_2$ are defined as:
  \(
    \FST \equiv \la p. p\,(\la x.\la y. x) \text{ and }
    \SND \equiv \la p. p\,(\la x.\la y. y)
  \).
\end{definition}

\begin{proposition}
  A key property of $ \PAIR $ is:
  \(
    \PAIR \, A \, B \equiv_\beta 
    \PAIR \, (\FST \, (\PAIR \, A \, B)) \, (\SND \, (\PAIR \, A \, B))
  \).
\end{proposition}

\subsubsection{\centering Lists}
Similarly to natural church encoding, lists are treated as a function where: \texttt{NIL} selects the base case, and \texttt{CONS} builds a new list by applying a function to the head and recursing on the tail. This way, operations on lists can be defined using functions alone.

\begin{definition}
  We define the empty list in the same spirit as \texttt{0} and \FALSE:
  \(
    \NIL \equiv \la f.\la z. z
  \)
  It represents the terminating element of the inductive definition of lists. To build non-empty lists we use:
  \(
    \texttt{CONS} \equiv \la h.\la t.\la f.\la z. f\,h\,(t\,f\,z)
  \)
  which binds a new element $h$ to a pre-existing list $t$ or to \NIL.
\end{definition}
\begin{definition}
  Some other relevant list operators are:
  \[
    \texttt{ISNIL} \equiv \la l. l\,(\la h.\la t. \FALSE)\,\TRUE
  \]
  \[
    \texttt{HEAD} \equiv \la l.\,((l \; (\la h.\, \la t.\, h))\; \texttt{NIL}) \qquad
    \texttt{TAIL} \equiv \la l.\la c.\la n.l\ (\la h.\la r.\la g.g\ h\ (r\ c))\ (\la c.n)\ (\la h.\la t.t)
  \]
\end{definition}



